---
title: "prop_scores_matching"
output: pdf_document
date: "2026-03-30"
---

# Loading in Required Packages 
```{r}
library(tidyverse)
library(dplyr)
library(optmatch)
library(ggplot2)
```

# Reading in Data Frame 
```{r}
clean_df <- read_csv("Data/clean_df.csv")
```

# EDA

Mainly Investigating NA and missingness patterns. 
```{r}
clean_df %>% 
  group_by(smoked_100_cigs) %>% 
  summarise(n = n())

clean_df %>% 
  group_by(born_in_US) %>% 
  summarise(n = n())

clean_df %>% 
  group_by(mod_activity_per_week) %>% 
  summarise(n = n())

clean_df %>% 
  group_by(vig_activity_per_week) %>% 
  summarise(n = n())

clean_df %>% 
  group_by(cig_per_day) %>% 
  summarise(n = n())
```

```{r}
table(clean_df$race_ethnicity_detailed)
```
```{r}
clean_df %>% 
  distinct() %>% 
  filter(!is.na(insecticide))
```

```{r}
clean_df %>% 
  group_by(active_duty_military) %>% 
  summarise(n = n())

table(clean_df$active_duty_military)

clean_df %>% 
  group_by(years_in_us) %>% 
  summarise(n = n())
```

```{r}
table(clean_df$years_in_us)
```

```{r}
clean_df %>% 
  group_by(hh_reference_gender) %>% 
  summarise(n = n())
```

```{r}
clean_df %>% 
  group_by(alcohol_avg_drinks_per_day) %>% 
  summarise(n = n())
```

```{r}
plot(
  clean_df$sleep_hours_weekday,
  clean_df$sleep_hours_weekend,
  main = "Sleep Hours: Weekday vs Weekend",
  xlab = "Sleep Hours Weekday",
  ylab = "Sleep Hours Weekend",
  pch = 19,
  col = rgb(0.2, 0.4, 0.6, 0.6)
)
```

```{r}
plot(
  clean_df$mod_activity_per_week,
  clean_df$sedentary_minutes,
  main = "Moderate Activity vs Sedentary Time",
  xlab = "Moderate Activity (per week)",
  ylab = "Sedentary Minutes",
  pch = 19,
  col = rgb(0.2, 0.4, 0.6, 0.6)
)
```



```{r}
na_prevalance <- clean_df %>% 
  is.na() %>% 
  colMeans() 

na_prevalance > 0.5
```


```{r}
na_counts <- sapply(clean_df, function(x) sum(is.na(x)))
na_counts 
```
```{r}
table(clean_df$insecticide)
```
```{r}

ggplot(clean_df, aes(x = race_ethnicity_detailed)) +
  geom_bar(fill = "#336699") +
  facet_wrap(~ insecticide) +
  labs(
    title = "Race Distribution by Treatment Group",
    x = "Race",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r}
ggplot(clean_df, aes(x = gender)) +
  geom_bar(fill = "#336699") +
  facet_wrap(~ insecticide) +
  labs(
    title = "Gender Distribution by Treatment Group",
    x = "Race",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
```{r}
# Create a summary data frame with means per group
means_df <- clean_df %>%
  group_by(insecticide) %>%
  summarise(mean_val = mean(systolic_avg, na.rm = TRUE))

ggplot(clean_df, aes(x = systolic_avg)) +
  geom_histogram(fill = "#336699") +
  geom_vline(
    data = means_df,
    aes(xintercept = mean_val),
    color = "red",
    linetype = "dashed",
    size = 1
  ) +
  facet_wrap(~ insecticide) +
  labs(
    title = "Blood Pressure by Treatment Group",
    x = "Blood Pressure (Systolic Average)",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
geom_text(
  data = means_df,
  aes(x = mean_val + 15, y = Inf, label = round(mean_val, 1)),
  vjust = 2,
  color = "black"
)
```

# Making SubjectID Numbers
```{r}
clean_df = clean_df %>% 
  mutate(subject_ID = `...1`)
```

# Complete Cases 

Number of Complete Cases for Required Covariates. 
```{r}
sum(complete.cases(clean_df[, c(
  "insecticide", "pulse_avg", "gender", "age_years",
  "race_ethnicity_detailed", "active_duty_military",
  "education_level", "marital_status", "household_size",
  "income_poverty_ratio", "diabetes_diagnosed",
  "sedentary_minutes", "sleep_hours_weekday",
  "smoked_100_cigs", "height_inches", "weight_pounds",
  "mod_activity_per_week"
)]))
```

Complete Cases Indices. 
```{r}
complete_cases <- complete.cases(clean_df[, c(
  "insecticide", "pulse_avg", "gender", "age_years",
  "race_ethnicity_detailed", "active_duty_military",
  "education_level", "marital_status", "household_size",
  "income_poverty_ratio", "diabetes_diagnosed",
  "sedentary_minutes", "sleep_hours_weekday",
  "smoked_100_cigs", "height_inches", "weight_pounds",
  "mod_activity_per_week"
)])
```

# Complete Cases DataFrame
```{r}
complete_cases_df <- clean_df[complete_cases, ]

table(complete_cases_df$insecticide)
```

# Standardized Differences Before Matching 
On Complete Cases Data. 

```{r}
table(complete_cases_df$race_ethnicity_detailed)

sum(is.na(clean_df$race_ethnicity))
sum(is.na(clean_df$race_ethnicity_detailed))
```


```{r}
complete_cases_df_encoded = complete_cases_df %>% 
  mutate(gender = ifelse(gender == "Female", 1, 0), # Female is 1, Male 0
         Mexican = ifelse(race_ethnicity_detailed == "Mexican American", 1, 0), 
         Asian = ifelse(race_ethnicity_detailed == "Non-Hispanic Asian", 1, 0),
         Black = ifelse(race_ethnicity_detailed == "Non-Hispanic Black", 1, 0), 
         White = ifelse(race_ethnicity_detailed == "Non-Hispanic White", 1, 0), 
         Hispanic = ifelse(race_ethnicity_detailed == "Other Hispanic", 1, 0), 
         Other = ifelse(race_ethnicity_detailed == "Other Race", 1, 0), 
         Less_9th_grade = ifelse(education_level == "<9th grade", 1, 0), 
         Some_High_School = ifelse(education_level == "9-11th grade", 1, 0), 
         High_School_Degree = ifelse(education_level == "High school/GED", 1, 0), 
         Some_College = ifelse(education_level == "Some college/AA", 1, 0), 
         College_Grad_or_More = ifelse(education_level == "College graduate or above", 1, 0), 
         Divorced = ifelse(marital_status == "Divorced", 1, 0), 
         Married = ifelse(marital_status == "Married", 1, 0), 
         Widowed = ifelse(marital_status == "Widowed", 1, 0)) 
```

```{r}
standardized_differences_df = complete_cases_df_encoded[, c(
  "insecticide", "pulse_avg", "gender", "age_years",
  "Mexican", "Asian", "Black", "White", "Hispanic", "Other", 
  "active_duty_military", 
  "Less_9th_grade", "Some_High_School", "High_School_Degree", "Some_College", "College_Grad_or_More",
  "Divorced", "Married", "Widowed", 
  "household_size",
  "income_poverty_ratio", "diabetes_diagnosed",
  "sedentary_minutes", "sleep_hours_weekday",
  "smoked_100_cigs", "height_inches", "weight_pounds",
  "mod_activity_per_week"
)]
```

```{r}
treatedmat=standardized_differences_df %>% 
  filter(insecticide == 1)
# Standardized differences before matching
controlmat.before=standardized_differences_df %>% 
  filter(insecticide == 0)

controlmean.before=apply(controlmat.before,2,mean,na.rm=TRUE);
treatmean=apply(treatedmat,2,mean,na.rm=TRUE);
treatvar=apply(treatedmat,2,var,na.rm=TRUE);
controlvar=apply(controlmat.before,2,var,na.rm=TRUE);
stand.diff.before=(treatmean-controlmean.before)/sqrt((treatvar+controlvar)/2);
```

# Imputation 

Mean for numeric variables and mode for categorical variables. 

```{r}
# Imputation 
# Function to calculate mode
get_mode <- function(x) {
  ux <- unique(x[!is.na(x)])
  ux[which.max(tabulate(match(x, ux)))]
}

# Copy your dataset (so you don't overwrite original)
imputed_df <- clean_df

# Loop through each column
for (col in names(imputed_df)) {
  
  # If numeric --> mean imputation
  if (is.numeric(imputed_df[[col]])) {
    mean_val <- mean(imputed_df[[col]], na.rm = TRUE)
    imputed_df[[col]][is.na(imputed_df[[col]])] <- mean_val
  }
  
  # If factor or character --> mode imputation
  else if (is.factor(imputed_df[[col]]) || is.character(imputed_df[[col]])) {
    mode_val <- get_mode(imputed_df[[col]])
    imputed_df[[col]][is.na(imputed_df[[col]])] <- mode_val
    
    # Optional: convert character --> factor (helps with glm)
    if (is.character(imputed_df[[col]])) {
      imputed_df[[col]] <- as.factor(imputed_df[[col]])
    }
  }
}
```

# Model (with Complete Cases)

```{r}
# propensity score model 
prop_score_mod <- glm(insecticide ~ pulse_avg + 
      gender + 
      age_years + 
      race_ethnicity_detailed + 
      active_duty_military + 
      education_level + 
      marital_status + 
      household_size + 
      income_poverty_ratio + 
      diabetes_diagnosed + 
      sedentary_minutes + 
      sleep_hours_weekday + 
      smoked_100_cigs + 
      height_inches + 
      weight_pounds + 
      mod_activity_per_week, 
      data = complete_cases_df, family = binomial,
      x = TRUE, y = TRUE)

prop_score_mod_encoded <- glm(insecticide ~ pulse_avg + 
      gender + 
      age_years + 
      Mexican + 
      Asian + 
      Black + 
      White + 
      Hispanic + 
      Other + 
      active_duty_military + 
      Less_9th_grade + 
      Some_High_School + High_School_Degree + 
      Some_College + 
      College_Grad_or_More + 
      Divorced + 
      Married + 
      Widowed +   
      household_size + 
      income_poverty_ratio + 
      diabetes_diagnosed + 
      sedentary_minutes + 
      sleep_hours_weekday + 
      smoked_100_cigs + 
      height_inches + 
      weight_pounds + 
      mod_activity_per_week, 
      data = complete_cases_df_encoded, family = binomial,
      x = TRUE, y = TRUE)
```

```{r}
summary(prop_score_mod)
```
```{r}
Z = prop_score_mod$y
logitps = predict(prop_score_mod)
length(logitps)
```


```{r}
Z_encoded = prop_score_mod_encoded$y
logitps_encoded = predict(prop_score_mod_encoded)
length(logitps_encoded)
```

# Checking Overlap with Model with Complete Cases
```{r}
plot(density(logitps[Z]), col = "red", xlim = c(0,1) , main = "Density Estimates of Propensity Scores")
lines(density(logitps[!Z]))
legend("topleft", c("Treated", "Control"), col = c("red", "black"), lty = c(1,1))
```

# Checking Overlap with all Data (using model with just complete cases)
```{r}
covariates <- c(
    "pulse_avg",
    "gender",
    "age_years",
    "race_ethnicity_detailed",
    "active_duty_military",
    "education_level",
    "marital_status",
    "household_size",
    "income_poverty_ratio",
    "diabetes_diagnosed",
    "sedentary_minutes",
    "sleep_hours_weekday",
    "smoked_100_cigs",
    "height_inches",
    "weight_pounds",
    "mod_activity_per_week"
  )
  
#without imputing, limited observations
treated <- clean_df[clean_df$insecticide == TRUE, c(covariates, "insecticide")]
control <- clean_df[clean_df$insecticide == FALSE, c(covariates, "insecticide")]
```

```{r}
treated_prop_scores <- predict(prop_score_mod, treated, type = "response")
hist(treated_prop_scores)

control_prop_scores <- predict(prop_score_mod, control, type = "response")
hist(control_prop_scores)
```

# Checking overlap with Imputed Data (with model with just complete cases)
```{r}
#prop score model with clean cases, impute covariates for predicted prop scores so everyone
#has a prop score: 

treated_imp <- imputed_df[imputed_df$insecticide == TRUE, c(covariates, "insecticide")]
control_imp <- imputed_df[imputed_df$insecticide == FALSE, c(covariates, "insecticide")]
```

```{r}
treated_prop_scores_imp <- predict(prop_score_mod, treated_imp, type = "response")
hist(treated_prop_scores_imp)

control_prop_scores_imp <- predict(prop_score_mod, control_imp, type = "response")
hist(control_prop_scores_imp)
```

```{r}
range(treated_prop_scores_imp)
range(control_prop_scores_imp)
```

```{r}
# report the fraction of treated and control individuals that
# have an estimated propensity score below 0.0001

cat("Treated Proportion with Propensity Scores < 0.0001", mean(treated_prop_scores_imp < 0.0001), "\n")
cat("Control Proportion with Propensity Scores < 0.0001",  mean(control_prop_scores_imp < 0.0001))

#nothing crazy
```

```{r}
#final prop score: 

prop_score_df <- imputed_df[, c(covariates, "insecticide")]
prop_scores <- predict(prop_score_mod, prop_score_df, type = "response")
```

```{r}
length(prop_scores)
length(logitps)
```





# Distance Matrix 

```{r}
distmat.logitps = outer(logitps[Z == 1],logitps[Z==0],"-")
distmat.logitps = abs(distmat.logitps)

rownames(distmat.logitps)= rownames(complete_cases_df)[Z == 1]
colnames(distmat.logitps)= rownames(complete_cases_df)[Z == 0]
```


```{r}
distmat.logitps_encoded = outer(logitps_encoded[Z_encoded == 1],logitps_encoded[Z_encoded==0],"-")
distmat.logitps_encoded = abs(distmat.logitps_encoded)

rownames(distmat.logitps_encoded)= rownames(complete_cases_df_encoded)[Z_encoded == 1]
colnames(distmat.logitps_encoded)= rownames(complete_cases_df_encoded)[Z_encoded == 0]
```

# Implementing Calipers on Distance Matrix 
```{r}
addcaliper=function(dmat,z,logitp,calipersd=.2,penalty=1000){
sd.logitp=sd(logitp)
adif=abs(outer(logitp[z==1],logitp[z==0],"-"))
adif=(adif-(calipersd*sd.logitp))*(adif>(calipersd*sd.logitp))
dmat=dmat+adif*penalty
dmat
}
```

Distance Matrix with Caliper
```{r}
distmat_caliper=addcaliper(distmat.logitps,Z,logitps,calipersd = 0.5)

distmat_caliper_encoded=addcaliper(distmat.logitps_encoded,Z_encoded,logitps_encoded,calipersd = 0.5)
```

# Matching

With non-calipered distance matrix (with model for only complete cases on only complete cases dataframe).
```{r}
noControls = 1
matchvec=pairmatch(distmat.logitps,controls=noControls,data=complete_cases_df)
```

```{r}
summary(matchvec)
```
Generating matrix for paired matches. 

```{r}
generating_matched_pair_mat <- function(matched_vec, df){
  matchvec.num = as.numeric(substr(matched_vec,start=3,stop=10))
  matchvec.num.notNA = matchvec.num[!is.na(matchvec.num)] #To remove individuals who didn't get matched.
  matchID <- unique(matchvec.num.notNA)
  
  I = length(matchID)
  
  matchedPairMat = matrix(0,I,4)
  colnames(matchedPairMat) = c("SubjectID (Treated)","SubjectID (Control)","PS (Treated)","PS (Control)")
  treatedSubjID = rownames(df)[Z==1]
  controlSubjID = rownames(df)[Z==0]
  
  treatedSubjID = as.integer(treatedSubjID)
  controlSubjID = as.integer(controlSubjID)
  
  for(i in 1:I) {
  subjectIDs = which(matchvec.num == matchID[i])
  matchedPairMat[i,"SubjectID (Treated)"] = subjectIDs[subjectIDs %in% treatedSubjID]
  matchedPairMat[i,"SubjectID (Control)"] = subjectIDs[subjectIDs %in% controlSubjID]
  matchedPairMat[i,"PS (Treated)"] = round(logitps[matchedPairMat[i,"SubjectID (Treated)"]],3)
  matchedPairMat[i,"PS (Control)"] = round(logitps[matchedPairMat[i,"SubjectID (Control)"]],3)
  }
  return(matchedPairMat)
}


df_matched_pairs_no_calip <- generating_matched_pair_mat(matchvec, complete_cases_df)

```

Matching with calipered distance and generating matrix. 
```{r}
matchvec_caliper=pairmatch(distmat_caliper,controls=noControls,data=complete_cases_df)

df_matched_pairs_calip <- generating_matched_pair_mat(matchvec_caliper, complete_cases_df)
```


Matching with encoded data. 
```{r}
matchvec_encoded=pairmatch(distmat.logitps_encoded,controls=noControls,data=complete_cases_df_encoded)

df_matched_pairs_encoded <- generating_matched_pair_mat(matchvec_encoded, complete_cases_df_encoded)
```

Matched with encoded data calipers. 
```{r}
matchvec_caliper_encoded=pairmatch(distmat_caliper_encoded,controls=noControls,data=complete_cases_df_encoded)

df_matched_pairs_calip_encoded <- generating_matched_pair_mat(matchvec_caliper_encoded, complete_cases_df_encoded)
```

# Standardized Differences After Matching 
```{r}
controlmat.after=standardized_differences_df[df_matched_pairs_encoded[,2],];
controlmean.after=apply(controlmat.after,2,mean);

stand.diff.after=(treatmean-controlmean.after)/sqrt((treatvar+controlvar)/2);

standBeforeAfter = cbind(stand.diff.before[-1],stand.diff.after[-1])

colnames(standBeforeAfter ) = c("Before Match (Standardized Diff)",
"After Match (Standardized Diff)")
knitr::kable(round(abs(standBeforeAfter),3), caption = "Differences in Covariates (Before and After)")
```

# Love Plot 
```{r}
abs.stand.diff.before=abs(stand.diff.before[-1])
abs.stand.diff.after=abs(stand.diff.after[-1])
covariates=names(stand.diff.before[-1])
plot.dataframe=data.frame(abs.stand.diff=c(abs.stand.diff.before,abs.stand.diff.after),covariates=rep(covariates,2),type=c(rep("Before",length(covariates)),rep("After",length(covariates))))
ggplot(plot.dataframe,aes(x=abs.stand.diff,y=covariates))+geom_point(size=5,aes(shape=factor(type)))+scale_shape_manual(values=c(4,1))+geom_vline(xintercept=c(.1,.2),lty=2)
```








