
#Reading / Libraries 

```{r}
library(tidyverse)
library(dplyr)
library(optmatch)
library(ggplot2)
library(forcats)

clean_df <- read_csv("Data/clean_df.csv")

clean_df = clean_df %>% 
  mutate(subject_ID = `...1`)
```

# MODEL with Imputed Data (and indictator of whether value is imputed)
```{r}
covariates <- c(
    "pulse_avg",
    "gender",
    "age_years",
    "race_ethnicity_detailed",
    "active_duty_military",
    "education_level",
    "marital_status",
    "household_size",
    "income_poverty_ratio",
    "diabetes_diagnosed",
    "sedentary_minutes",
    "sleep_hours_weekday",
    "smoked_100_cigs",
    "height_inches",
    "weight_pounds",
    "mod_activity_per_week")
 
variables_of_interest <- c(covariates, "mean_bp", "insecticide", "hearing_cond")


na_counts <- sapply(clean_df[, variables_of_interest], function(x) sum(is.na(x)))
na_counts #na_counts among variables of interest  
```
```{r}
clean_df <- clean_df %>% 
  mutate(gender = ifelse(gender == "Female", 1, 0), # Female is 1, Male 0
         Mexican = ifelse(race_ethnicity_detailed == "Mexican American", 1, 0), 
         Asian = ifelse(race_ethnicity_detailed == "Non-Hispanic Asian", 1, 0),
         Black = ifelse(race_ethnicity_detailed == "Non-Hispanic Black", 1, 0), 
         White = ifelse(race_ethnicity_detailed == "Non-Hispanic White", 1, 0), 
         Hispanic = ifelse(race_ethnicity_detailed == "Other Hispanic", 1, 0), 
         Other = ifelse(race_ethnicity_detailed == "Other Race", 1, 0)) %>% 
  
  mutate(deaf = ifelse(hearing_cond == "Deaf", 1, 0), 
           excellent = ifelse(hearing_cond == "Excellent", 1, 0), 
           good = ifelse(hearing_cond == "Good", 1, 0), 
           little_trouble = ifelse(hearing_cond == "A Little Trouble", 1, 0), 
           lot_trouble = ifelse(hearing_cond == "Lot of Trouble", 1, 0), 
           moderate_trouble = ifelse(hearing_cond == "Moderate Trouble", 1, 0), 
           hearing_cond = factor(hearing_cond, levels = c("Excellent", 
                                                          "Good", 
                                                          "A Little Trouble", 
                                                          "Moderate Trouble", 
                                                          "Lot of Trouble", 
                                                          "Deaf")))

```

```{r}
final_imputed_df <- clean_df

covariates_to_impute <- c(
    "active_duty_military",
    "education_level",
    "marital_status",
    "income_poverty_ratio",
    "diabetes_diagnosed",
    "sedentary_minutes",
    "sleep_hours_weekday",
    "smoked_100_cigs",
    "height_inches",
    "weight_pounds",
    "mod_activity_per_week"
  )
  


for (col in covariates_to_impute) {
  
  # 1. get column
  x <- final_imputed_df[[col]]
  
  # 2. create indicator 
  final_imputed_df[[paste0(col, "_imputed")]] <- ifelse(is.na(x), 1, 0)
  
  # 3. compute imputation value
  if (is.numeric(x)) {
    impute_val <- mean(x, na.rm = TRUE)
  } else {
    impute_val <- names(sort(table(x), decreasing = TRUE))[1]
  }
  
  # 4. replace NA values
  x[is.na(x)] <- impute_val
  
  if (is.character(x)) {
    x <- as.factor(x)
  }
  
  # 5. put column back
  final_imputed_df[[col]] <- x
}

```

```{r}
# re-do encoding for columns that needed imputation 
final_imputed_df = final_imputed_df %>% 
  mutate(Less_9th_grade = ifelse(education_level == "<9th grade", 1, 0), 
         Some_High_School = ifelse(education_level == "9-11th grade", 1, 0), 
         High_School_Degree = ifelse(education_level == "High school/GED", 1, 0), 
         Some_College = ifelse(education_level == "Some college/AA", 1, 0), 
         College_Grad_or_More = ifelse(education_level == "College graduate or above", 1, 0), 
         Prev_Married = ifelse(marital_status == "Previously Married", 1, 0), 
         Married = ifelse(marital_status == "Married", 1, 0), 
         Never_Married = ifelse(marital_status == "Never Married", 1, 0)) 
```

```{r}
final_imputed_df
```

```{r}
prop_score_mod_imputed <- glm(insecticide ~ pulse_avg +
      gender + 
      age_years + 
      Mexican + Asian + Black + White + Hispanic + 
        Other +
      active_duty_military + active_duty_military_imputed +
      Less_9th_grade + Some_High_School + High_School_Degree + 
        Some_College + College_Grad_or_More + education_level_imputed +
      Prev_Married + Married + Never_Married + marital_status_imputed +
      household_size + 
      income_poverty_ratio + income_poverty_ratio_imputed +
      diabetes_diagnosed + diabetes_diagnosed_imputed +
      sedentary_minutes + sedentary_minutes_imputed +
      sleep_hours_weekday + sleep_hours_weekday_imputed +
      smoked_100_cigs + smoked_100_cigs_imputed +
      height_inches + height_inches_imputed +
      weight_pounds + weight_pounds_imputed +
      mod_activity_per_week + mod_activity_per_week_imputed,
      data = final_imputed_df, family = binomial,
      x = TRUE, y = TRUE)
```

```{r}
summary(prop_score_mod_imputed)
```


# Check Overlap


```{r}
prop_scores <- predict(prop_score_mod_imputed, final_imputed_df, type = "response")
treated_prop_scores <- prop_scores[final_imputed_df$insecticide == TRUE]
control_prop_scores <- prop_scores[final_imputed_df$insecticide == FALSE]

```



```{r}
par(mfrow = c(1, 2))

hist(treated_prop_scores, xlab = "Propensity Scores", main = "Propensity Scores for Treated")

hist(control_prop_scores, xlab = "Propensity Scores", main = "Propensity Scores for Control")
```



```{r}
# report the fraction of treated and control individuals that
# have an estimated propensity score below 0.0001

cat("Treated Proportion with Propensity Scores < 0.0001", mean(treated_prop_scores < 0.0001), "\n")
cat("Control Proportion with Propensity Scores < 0.0001",  mean(control_prop_scores < 0.0001))

#nothing crazy
```

# Distance Matrix + Caliper 

```{r}
Z = prop_score_mod_imputed$y
length(prop_scores)

logitps = predict(prop_score_mod_imputed)
length(logitps)

```


```{r}
distmat.logitps = outer(logitps[Z == 1],logitps[Z==0],"-")
distmat.logitps = abs(distmat.logitps)

rownames(distmat.logitps)= rownames(final_imputed_df)[Z == 1]
colnames(distmat.logitps)= rownames(final_imputed_df)[Z == 0]
```


```{r}
addcaliper = function(dmat,z,logitp,calipersd=.2,penalty=1000){
  sd.logitp=sd(logitp)
  adif=abs(outer(logitp[z==1],logitp[z==0],"-"))
  adif=(adif-(calipersd*sd.logitp))*(adif>(calipersd*sd.logitp))
  dmat=dmat+adif*penalty
  dmat
}
```

```{r}
distmat_caliper = addcaliper(distmat.logitps, Z, logitps, calipersd = 0.5)
```

# Matching 

```{r}
noControls = 1

matchvec_caliper=pairmatch(distmat_caliper,controls=noControls,data=final_imputed_df)
summary(matchvec_caliper)
```

```{r}
generating_matched_pair_mat <- function(matched_vec, df){
  matchvec.num = as.numeric(substr(matched_vec,start=3,stop=10))
  matchvec.num.notNA = matchvec.num[!is.na(matchvec.num)] #To remove individuals who didn't get matched.
  matchID <- unique(matchvec.num.notNA)
  
  I = length(matchID)
  
  matchedPairMat = matrix(0,I,4)
  colnames(matchedPairMat) = c("SubjectID (Treated)","SubjectID (Control)","PS (Treated)","PS (Control)")
  treatedSubjID = rownames(df)[Z==1]
  controlSubjID = rownames(df)[Z==0]
  
  treatedSubjID = as.integer(treatedSubjID)
  controlSubjID = as.integer(controlSubjID)
  
  for(i in 1:I) {
  subjectIDs = which(matchvec.num == matchID[i])
  matchedPairMat[i,"SubjectID (Treated)"] = subjectIDs[subjectIDs %in% treatedSubjID]
  matchedPairMat[i,"SubjectID (Control)"] = subjectIDs[subjectIDs %in% controlSubjID]
  matchedPairMat[i,"PS (Treated)"] = round(logitps[matchedPairMat[i,"SubjectID (Treated)"]],3)
  matchedPairMat[i,"PS (Control)"] = round(logitps[matchedPairMat[i,"SubjectID (Control)"]],3)
  }
  return(matchedPairMat)
}
```

```{r}
df_matched_pairs_calip <- generating_matched_pair_mat(matchvec_caliper, final_imputed_df)
```

# Standarized Differences 

```{r}
standardized_differences_df = final_imputed_df[, c(
  "insecticide", "pulse_avg", "gender", "age_years",
  "Mexican", "Asian", "Black", "White", "Hispanic", "Other", 
  "active_duty_military", 
  "Less_9th_grade", "Some_High_School", "High_School_Degree", "Some_College", "College_Grad_or_More",
  "Prev_Married", "Married", "Never_Married", 
  "household_size",
  "income_poverty_ratio", "diabetes_diagnosed",
  "sedentary_minutes", "sleep_hours_weekday",
  "smoked_100_cigs", "height_inches", "weight_pounds",
  "mod_activity_per_week"
)]
```


```{r}
treatedmat=standardized_differences_df %>% 
  filter(insecticide == 1)
# Standardized differences before matching
controlmat.before=standardized_differences_df %>% 
  filter(insecticide == 0)

controlmean.before=apply(controlmat.before,2,mean,na.rm=TRUE);
treatmean=apply(treatedmat,2,mean,na.rm=TRUE);
treatvar=apply(treatedmat,2,var,na.rm=TRUE);
controlvar=apply(controlmat.before,2,var,na.rm=TRUE);
stand.diff.before=(treatmean-controlmean.before)/sqrt((treatvar+controlvar)/2);
```

```{r}
controlmat.after=standardized_differences_df[df_matched_pairs_calip[,2],];
controlmean.after=apply(controlmat.after,2,mean);

stand.diff.after=(treatmean-controlmean.after)/sqrt((treatvar+controlvar)/2);

standBeforeAfter = cbind(stand.diff.before[-1],stand.diff.after[-1])

colnames(standBeforeAfter ) = c("Before Match (Standardized Diff)",
"After Match (Standardized Diff)")
knitr::kable(round(abs(standBeforeAfter),3), caption = "Differences in Covariates (Before and After)")
```


# Love Plot 
```{r}
abs.stand.diff.before=abs(stand.diff.before[-1])
abs.stand.diff.after=abs(stand.diff.after[-1])
covariates=names(stand.diff.before[-1])
plot.dataframe=data.frame(abs.stand.diff=c(abs.stand.diff.before,abs.stand.diff.after),covariates=rep(covariates,2),type=c(rep("Before",length(covariates)),rep("After",length(covariates))))




plot.dataframe$covariates <- fct_recode(plot.dataframe$covariates,
  "Weight (lbs)" = "weight_pounds",
  "Education (Some HS)" = "Some_High_School",
  "Education (Some College)" = "Some_College", 
  "Education (<9th)" = "Less_9th_grade", 
  "Education (HS)" = "High_School_Degree",
  "Education (College or More)" = "College_Grad_or_More",
  "Smoked 100 cigs/daily" = "smoked_100_cigs", 
  "Sleep Time (hrs)" = "sleep_hours_weekday", 
  "Sedentary Time (mins)" = "sedentary_minutes", 
  "Average Pulse" = "pulse_avg", 
  "Race (Other)" = "Other", 
  "Race (White)" = "White", 
  "Race (Mexican)" = "Mexican", 
  "Race (Black)" = "Black", 
  "Race (Asian)" = "Asian", 
  "Race (Hispanic)" = "Hispanic",
  "Marital Status (Previously Married)" = "Prev_Married", 
  "Marital Status (Married)" = "Married", 
  "Marital Status (Never Married)" = "Never_Married", 
  "Moderate Physical Activity (/week)" = "mod_activity_per_week", 
  "Income Poverty Ratio" = "income_poverty_ratio", 
  "Household Size" = "household_size", 
  "Height (in)" = "height_inches", 
  "Gender" = "gender", 
  "Diabetes Diagnosis" = "diabetes_diagnosed", 
  "Age (yrs)" = "age_years", 
  "Active Duty Military" = "active_duty_military"
)

ggplot(plot.dataframe,aes(x=abs.stand.diff,y=covariates))+
  geom_point(size=5,aes(shape=factor(type)))+
  scale_shape_manual(values=c(4,1))+geom_vline(xintercept=c(.1,.2),lty=2) + 
  labs(shape = "Before or After Matching", 
       title = "Standardized Differences for Imputed Data", 
       y = "Covariates", 
       x = "Absolute Standardized Differences") + 
  theme_bw()
```

# Hearing Outcome: 

Check if there's a difference between the matched pairs in terms of hearing condition

```{r}
conds <- c("excellent", 
  "good", 
  "little_trouble", 
  "moderate_trouble",
  "lot_trouble", 
  "deaf"
)

get_info <- function(full_df, matched_pairs, response) {
  treated = full_df[matched_pairs[,"SubjectID (Treated)"], response]
  treated_response = treated[[response]]
  control = full_df[matched_pairs[,"SubjectID (Control)"], response]
  control_response = control[[response]]
  
  print(response)
  print(paste('control: ', sum(control_response)))
  print(paste('treated: ', sum(treated_response)))

}

# for (cond in conds) {
#   get_info(final_imputed_df, df_matched_pairs_calip, cond)
# }
```

```{r}
wilcoxon <- function(full_df, matched_pairs, response) {
  treated = full_df[matched_pairs[,"SubjectID (Treated)"], response]
  treated_response = treated[[response]]
  control = full_df[matched_pairs[,"SubjectID (Control)"], response]
  control_response = control[[response]]

  
  print(wilcox.test(treated_response, control_response, 
              paired=TRUE, conf.int=TRUE))
}


ttest <- function(full_df, matched_pairs, response) {
  treated = full_df[matched_pairs[,"SubjectID (Treated)"], response]
  treated_response = treated[[response]]
  control = full_df[matched_pairs[,"SubjectID (Control)"], response]
  control_response = control[[response]]

  print(t.test(treated_response, control_response, paired = TRUE, alternative = "two.sided"))
}

mcnemar <- function(full_df, matched_pairs, response) {
  treated = full_df[matched_pairs[,"SubjectID (Treated)"], response]
  t = treated[[response]]
  control = full_df[matched_pairs[,"SubjectID (Control)"], response]
  c = control[[response]]
  
  df <- data.frame(c, t) #find when 0 -> 1, etc 
  
  #print(df)
  
  table <-
  matrix(c(
    sum((c == 1) & (t == 1)),
    sum((c == 1) & (t == 0)),
    sum((c == 0) & (t == 1)),
    sum((c == 0) & (t == 0))
  ),
  nrow = 2
)
  
  print(table)
  print(mcnemar.test(table))
}
```


```{r}
for (cond in conds) {
  print(cond)
  mcnemar(final_imputed_df, df_matched_pairs_calip, cond)
  #wilcoxon(final_imputed_df, df_matched_pairs_calip, cond)
  #ttest(final_imputed_df, df_matched_pairs_calip, cond)
  

}
```
Excellent has a p-val below 0.05 

Wilcox has an error with deaf and lot of trouble, too few observations. 


# Blood Pressure Outcome Tests 

```{r}
for (cond in conds) {
  print(response)
  wilcoxon(final_imputed_df, df_matched_pairs_calip, response)
  ttest(final_imputed_df, df_matched_pairs_calip, response)
}
```

# Senstivity Analysis
```{r}
# is it correct to put diff >= for s1?? 

sens.analysis.signedrank=function(diff,Gamma){
  rk=rank(abs(diff));
  s1=1*(diff>=0);
  s2=1*(diff<0);
  W=sum(s1*rk);
  Eplus=sum((s1+s2)*rk*Gamma)/(1+Gamma);
  Eminus=sum((s1+s2)*rk)/(1+Gamma);
  V=sum((s1+s2)*rk*rk*Gamma)/((1+Gamma)^2);
  Dplus=(W-Eplus)/sqrt(V);
  Dminus=(W-Eminus)/sqrt(V);
  list(lowerbound=1-pnorm(Dminus),upperbound=1-pnorm(Dplus));
}
```


```{r}
treated = complete_cases_df[df_matched_pairs_calip_complete[,"SubjectID (Treated)"], "excellent"]
treated_response = treated[["excellent"]]
control = complete_cases_df[df_matched_pairs_calip_complete[,"SubjectID (Control)"], "excellent"]
control_response = control[["excellent"]]

difference = treated_response - control_response
```

```{r}
Gamvec = seq(1, 2, 0.1)
tt = matrix(0, length(Gamvec), 3)
tt[,1] = Gamvec
for(i in 1:length(Gamvec))
{
  a = sens.analysis.signedrank(difference, Gamvec[i])
  tt[i,-1] = c(a$lowerbound, a$upperbound)
}

colnames(tt) = c("Gamma", "Lower Bound", "Upper Bound")
knitr::kable(round(tt,3), caption = "Gamma and Bounds")
```


